---
title: "Statistical Inference Course Project"
author: "Daria Stepanyan"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```

## Synopsys

This report consists of 2 parts. First one is dedicated to investigation of the the exponential distribution and its comparison with the Central Limit Theorem. Second one contains analysis of ToothGrowth data in the R datasets package.

## Exponential distribution

### Simulation

For the purpose of this investigation we run a 1000 simulations of 40 exponentials with lambda = .2, taking their mean

```{r param}
lambda = 0.2
n = 40
nsim = 1000
mns = NULL
set.seed(2019)
for (i in 1 : nsim) {
        mns = c(mns, mean(rexp(n, lambda)))
}
```

Using the fact that mean fucntion is linear, we can calculate sample mean:
```{r sm}
mean(mns)
```

which is quite close to a theoretized value of 1/lambda, or `r 1/lambda`

We can also calculate variance values
```{r vars}
setNames(data.frame(matrix(data = c(1/(lambda^2)/n, var(mns)), 
        ncol = 2, nrow = 1)), 
        c("theoretical variance", "sample variance"))
```

CLT tells us that assuming the variables are iid the distribution of means of that variable will approximate normal distribution - even if original values are not normally distributed - and its mean will estimate a population mean. Let's see it on the plot.

First we create a simulated means distribution plot

```{r meanplot}
bw = .1
library(ggplot2) 
library(gridExtra)

## plot a histogram of simulation means
plotmean <- ggplot(mapping = aes(mns)) + 
        geom_histogram(alpha=0.3, fill='white', colour='black', binwidth = bw)
## add a theoretical mean to a plot (dashed red line)
plotmean <- plotmean + geom_vline(xintercept = 1/lambda, 
        linetype="dashed", color = "red")
## add the sample mean 
plotmean <- plotmean + geom_vline(xintercept = mean(mns), color = "blue")
## add titles
plotmean <- plotmean + labs(title = "Distribution of sample means of exponential function (1000 simulations)",
        x="sample means")

## add a normal distribution curve scaled to the sample mean distribution
tmean = 1/lambda
tsd = sqrt(1/(lambda^2)/n)

smean = mean(mns)
ssd = sd(mns)

plotnorm <- plotmean + 
        stat_function (fun = function(x, mean, sd, n){
                n * dnorm(x = x, mean = mean, sd = sd) * bw
                }, 
                args = c(mean = tmean, sd = tsd, n = nsim), color = "red") +
        stat_function (fun = function(x, mean, sd, n){
                n * dnorm(x = x, mean = mean, sd = sd) * bw
                }, 
                args = c(mean = smean, sd = ssd, n = nsim), color = "blue")

```

For truly seeing CLT in action we compare the distribution we got for plain exponentials obtained in `r nsim` simulations

```{r randsim}
randsam <- rexp(nsim, lambda)

plotrand <- ggplot(mapping = aes(randsam)) +
                geom_histogram (alpha=0.3, fill='white', 
                        colour='black', binwidth = bw) +
        labs(title = "Distribution of exponential function values (1000 simulations)", x="values") 

## adding the scale for better redability
plotrand <- plotrand + xlim(0,10)
plotnorm <- plotnorm + xlim(0,10)

## plot 2 distributions
grid.arrange(plotrand, plotnorm, nrow = 2)
```

We can see that even though the underlying data is not normally distributed, the distribution of sample means approximate the cassic bell curve of a normal distribution.

## 