---
title: "Analysis of the Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Daria Stepanyan"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, cache = TRUE)
```

## Synopsys

This report contains exploratory analysis of the tooth growth data and investigates statistical significance of reported effects.


## Summary of data

```{r data}
data(ToothGrowth)
str(ToothGrowth)
```

Dataset "ToothGrowth" coming with R contains data on the response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC).

```{r}
summary(ToothGrowth)
```

There don't seem to be any missing values and the values of len vary quite a lot. Let's plot the data to see if there are any obvious patterns

```{r groupplot, fig.height=3}
library(ggplot2) 

qplot(supp,len,data=ToothGrowth, facets=~dose, 
        main="ToothGrowth data: length vs supplement, given the dose",
        xlab="Supplement", ylab="Tooth length") + 
        geom_boxplot(aes(fill = supp))
```

For the smaller dosages orange juice seems to have more effect on tooth growth than vitamin C, but with the dosage increased to 2 mg the associated tooth growth is very similar.

## Assumptions

Since we don't know the underlying population characteristics we will have to make several assumtions:

* Population data is normally distributed
* The variables must be independent and identically distributed (i.i.d.).
* Variances of tooth growth are different when using different supplement and dosage

## Hypothesis

Let the H$_0$ be that the average tooth growth in a given dosage for both supplements are equal, and the alternative H$_1$ be that orange juice is more effective in smaller dosages (0.5, 1). 

For the dosage of 2 mg we'll perform 2-sided test with alternative hypothesis H$_1$: mu0 <> mu1

```{r}
table(ToothGrowth$supp, ToothGrowth$dose)
```

Each experiment contains exactly 10 observations, so t-test will be appropriate to use.

Let's also assume an acceptable alpha level at 0.05.

```{r setalpha}
alpha = 0.05
```

And run the t-tests for different dosages. In an essense we want to see if the difference in means of OJ experiments - VC experiments is greater than zero with 95% confidence.

```{r ttests}
tl <- split(ToothGrowth, ToothGrowth$dose)

testres = NULL

for (i in c(1:2)) {
        testres[[i]] <- with(tl[[i]], 
                t.test(len[supp == "OJ"], len[supp == "VC"],
                alternative = "g"))
}

testres[[3]] <- with(tl[[3]], 
                t.test(len[supp == "OJ"], len[supp == "VC"],
                alternative = "two"))
```

Now we can arrange test results into a table and determine the hypothesis status

```{r test-results}

## setting up a table for results
results <- setNames(data.frame(matrix(ncol = 7, nrow = 0)), 
        c("supp1", "supp2", "dose", "confint1", 
                "confint2", "p-value", "status"))

## creating a function to determine whether null hypothesis is rejected based on test results based on whether confidence interval contains 0, and p-value
hypo <- function (t) {
        outZero <- sign(prod(t$conf.int))
        if ((outZero != 1) || (t$p.value > alpha)) {
                "fail to reject"
        } else {
                "reject"
        }
}

## filling the data
for (i in c(1:3)) {
        results[i,"supp1"] <- "OJ"
        results[i,"supp2"] <- "VC"
        results[i,"dose"] <- tl[[i]]$dose[1]
        results[i,"confint1"] <- testres[[i]]$conf.int[1]
        results[i,"confint2"] <- testres[[i]]$conf.int[2]
        results[i,"p-value"] <- testres[[i]]$p.value
        results[i,"status"] <- hypo(testres[[i]])
}

results
```

## Conclusions
As a result of t-tests performed on the data we can colnclude that:
* For the smaller dosages (0.5, 1 mg) the tooth growing effect of orange juice is greater than that of vitamin C with p-values of `r results["p-value",1]` and `r results["p-value",2]` respectively
* for the bigger dosage (2 mg) we failed to reject the hypothesis that both supplrements have similar effect, and there's no reason to assume that effects of vitamin C and orange juice vary

### Environment data
```{r}
sessionInfo()
```